Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}

### 2014-05-10 Manuscript/template.tex
ENH:  Submit to Neuroinformatics.

%% This is file `elsarticle-template-1-num.tex',
%%
%% Copyright 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%%
%% $Id: elsarticle-template-1-num.tex 149 2009-10-08 05:01:15Z rishi $
%% $URL: http://lenova.river-valley.com/svn/elsbst/trunk/elsarticle-template-1-num.tex $
%%

\documentclass[preprint,authoryear,review,12pt]{elsarticle}
%\documentclass[final,5p,times,twocolumn]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,two column; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

\usepackage{color,hyperref,bm}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{lscape}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{ulem}
\usepackage{setspace}
\usepackage{listings}
\usepackage{float}
\usepackage{listings}
\usepackage{color,colortbl}
\usepackage{rccol}
\usepackage{pbox}
%\usepackage[table]{xcolor}

    \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
    \definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
    \definecolor{listnumbers}{gray}{0.65}
    \definecolor{listlightgray}{gray}{0.955}
    \definecolor{listwhite}{gray}{1.0}
    \definecolor{lightcyan}{rgb}{0.88,1,1}

\newcommand{\lstsetcpplong}
{
\lstset{frame = tb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false,
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=,
        floatplacement=!h,
        caption={\small \baselineskip 12pt DiReCT long command line menu which is invoked using the `{\ttfamily {-}{-}help}' option.  The short command line menu is obtained by typing `{\ttfamily {-}h}'},
        captionpos=b,
        label=listing:long
        }
}

\floatstyle{plain}
\newfloat{command}{thp}{lop}
\floatname{command}{Command}

%\usepackage[nomarkers,notablist]{endfloat}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
% \usepackage{amsthm}

 \usepackage{makecell}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%
%% \biboptions{comma,round}

% \biboptions{}

\providecommand{\OO}[1]{\operatorname{O}\bigl(#1\bigr)}

@@

\long\def\symbolfootnote[#1]#2{\begingroup%
\def\thefootnote{\fnsymbol{footnote}}\footnote[#1]{#2}\endgroup}

\journal{NeuroImage}

\begin{document}

\begin{frontmatter}

\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}}

\author[label1]{Nicholas J.~Tustison
  \fnref{label0}}
  \fntext[label0]{\scriptsize Corresponding author:  PO Box 801339, Charlottesville, VA 22908; T:  434-924-7730; email address:  ntustison@virginia.edu.
  }
\author[label2]{K.~L.~Shrinidhi}
\author[label1]{Max Wintermark}
\author[label1]{Christopher R.~Durst}
\author[label2]{Benjamin M.~Kandel}
\author[label2]{James C.~Gee}
\author[label3]{Murray C.~Grossman}
\author[label2]{Brian B.~Avants}
\address[label1]{Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA}
\address[label2]{Penn Image Computing and Science Laboratory, 
                 Department of Radiology, University of Pennsylvania,
                Philadelphia, PA}
\address[label3]{Department of Neurology, University of Pennsylvania,
                Philadelphia, PA}

%\maketitle

\begin{abstract} 
Segmenting and quantifying gliomas from MRI is an important task for diagnosis, planning intervention, 
and for tracking tumor changes over time.
However, this task is complicated by the lack
of prior knowledge concerning tumor location, spatial extent, shape,
possible displacement of normal tissue, and intensity signature.
%These factors limit the applicability of techniques with established
%performance in related tasks such as normal brain tissue
%segmentation.  
To accommodate such complications, we introduce a framework for supervised
segmentation based on multiple modality intensity, geometry, and asymmetry feature sets.  These features drive a supervised whole-brain and tumor segmentation
approach based on random forest-derived probabilities.  The
asymmetry-related features (based on optimal symmetric multimodal 
templates) demonstrate excellent discriminative 
properties within this framework. We also gain performance by  
generating probability maps from random forest models and using these maps 
for a refining Markov random field regularized probabilistic segmentation.  
This strategy allows us to interface the supervised learning capabilities 
of the random forest model with regularized probabilistic segmentation
using the recently developed \textit{ANTsR} 
package---a comprehensive statistical and visualization interface between the
popular Advanced Normalization Tools (ANTs) and the \textit{R}
statistical project. 
%\textit{ANTsR} integrates proven image analysis methods
%for image registration, segmentation, bias correction, and
%template building with \textit{R}'s statistical and machine
%learning techniques and state-of-the-art visualization possibilities.  
The reported algorithmic framework was the
top-performing entry in the MICCAI 2013 Multimodal Brain Tumor
Segmentation challenge.  The challenge data were widely varying 
consisting of both high-grade and low-grade glioma tumor 
four-modality MRI from five different institutions.
Average Dice overlap measures for the final algorithmic assessment
were 0.87, 0.78, and 0.74 for ``complete'', ``core'', and ``enhanced'' 
tumor components, respectively.
%and, to our knowledge, is the only entry for
%both challenge years (2012 \& 2013) which has been made publicly
%available.
\end{abstract}

\begin{keyword}
advanced normalization tools \sep BRATS \sep brain tumor segmentation 
\sep \textit{R} project for statistical computing
%% keywords here, in the form: keyword \sep keyword
\end{keyword}

\end{frontmatter}
%\linenumbers

%
%
\newpage

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

%%
%% Start line numbering here if you want
%%
% \linenumbers

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

%% \section{}
%% \label{}

%% References
%%
%% Following citation commands can be used in the body text:
%% Usage of \cite is as follows:
%%   \citep{key}          ==>>  [#]
%%   \cite[chap. 2]{key} ==>>  [#, chap. 2]
%%   \citet{key}         ==>>  Author [#]

%% main text

%\input{intro}
%
%\input{methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Introduction
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
Given the complexity of tumor growth and appearance and the need
for precise volumetric measurements for tumor characterization, 
much research has been invested in computational methods for 
automated segmentation of tumor regions in MR images.  Several 
approaches have been previously proposed in the literature as
detailed in recent reviews \citep{angelini2007,bauer2013}.  
Commensurable algorithmic evaluation, however, is
extremely problematic due to widely varying performance assessments
described in the corresponding publications, lack of publicly available 
evaluation data, and private algorithmic instantiations.  
In response to these issues, the Multimodal Brain Tumor Segmentation 
(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
the auspices of the Medical Image Computing and Computer Assisted 
Intervention Society in association with their annual international 
conference \citep{menze2014}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Materials and Methods
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}
  \centering
  \begin{tabular}{cccc}
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
  (a) & (b) &
  (c) & (d) \\
  \end{tabular}
  \caption{Induced bilateral asymmetry due to tumor presence causing
  distortion of the plane of symmetry.  Shown are mid-axial slices of
  one of the BRATS 2013 training data 
  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
  (d) T2.  }
  \label{fig:asymmetry}
\end{figure}

\begin{figure*}
    \centerline{\includegraphics[width=130mm]{templateKirby.pdf}}
  \caption{Multivariate symmetric template created from the MMRR 
           data \citep{landman2011}.  Of the seven modalities 
           comprising the set of study acquisitions, we illustrate the
           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
           Although DWI-based images were not included in the challenge data, 
           such images have shown discriminative potential \citep{price2003,cha2005} 
           warranting investigation in our future work.  
           The optimal transformation and averaging of the individual 
           subject images result in the asymmetric template represented at 
           the top of the middle column.  A horizontal reflection, $T_R$,
           perpendicular to the mid-sagittal plane  resulted in the 
           contralateral counterpart represented
           at the bottom of the second column.  The final template seen on
           the right is a result of repeating the template construction using
           the two asymmetric templates as input.
          }
  \label{fig:symmetrictemplates}
\end{figure*}

@@

\begin{figure}
  \centering
%    \includegraphics[width=80mm]{asymmetryFeatureImages.pdf}
    \includegraphics[width=120mm]{asymmetryFeatureImages.pdf}
  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.  $T_R$ denotes a horizontal reflection
  perpendicular to the mid-sagittal plane.
  Feature images used are specified by the dashed box.
Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
          }
  \label{fig:asymmetryFeatures}
\end{figure}

@@

\begin{figure*}
\centering
  \makebox[0.9\textwidth][c]{
  \includegraphics[width=150mm]{featureImages.pdf}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  final assessment data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Similarly, we 
           calculate the voxelwise difference image between the
           T1 and T1C image.  
           of 
           Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   This mapping is also used to produce the log Jacobian image
           and warped template distance mask.  Additionally, the 
           (T1 - T1C) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

@@

\begin{figure}
  \centering
   \vspace{-15mm}
  \makebox[0.9\textwidth][c]{
    \includegraphics[width=150mm]{pipelineTestingTraining.pdf}
    }
  \caption{Diagrammatic workflow for the proposed RF model training (left) and prediction
  (right).  For training, the feature images are first generated from the set of input 
  training multimodal MRI and symmetric multivariate template. The set of features images are 
  use to create the RF model for the first stage.  This first stage RF model is then applied
  to the set of input training data to yield spatial priors for seeding the generation of the
  second stage feature images.  These feature images are then used to create the second
  stage RF model.  For prediction of a single-subject segmentation, a similar scenario 
  is applied whereby the generated feature images are used as input to the RF models the
  second of which produces an initial segmentation estimate.  This estimate is then refined
  using a series of heuristically-derived binary morphological operations resulting in the 
  final labeled brain.
  }
  \label{fig:pipeline}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tabular}{c}
    \includegraphics[width=88mm]{gmm_map_mrf.pdf}
  \end{tabular}
  \caption{Visual comparison between the first and second RF stages
  on the evaluation {\tt BRATS\_HG0001} data set.
  Although the GMM intensity modeling does provide certain 
  discriminative benefit (see Figures \ref{fig:hgimportance}
  and \ref{fig:lgimportance}), the use of the RF-derived posteriors 
  coupled with the described MAP-MRF segmentation framework enhances 
  the accuracy of the intensity-based modeling features.    
  }
  \label{fig:stageComparison}
\end{figure}

@@

Additionally, the {\tt randomForest} package provides  measurements 
for determining the importance of chosen features when constructing the model.  
This aids in potential feature pruning or intuiting model behavior.  As mentioned
earlier, we provide the \textit{R} script {\tt plotVariableImportance.R} to render
one such quantity use to assess RF feature importance denoted as {\tt MeanDecreaseAccuracy}.  
During model construction
(specifically the out-of-bag error calculation stage), the decrease in prediction accuracy
with the omission of a single feature or variable is tracked and averaged.  Thus,
those features which have the greatest decrease in mean accuracy are considered
to be the most discriminative.
%\footnote{
%There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.  Correctives have been proposed but we ignore these issues for this particular application.
%}
In this work, we do not use these measurements for feature pruning.  However,
we plot them in the Results section (see Figures \ref{fig:hgimportance} and
\ref{fig:lgimportance}) as they demonstrate the relative importance of our
selected features including that of the proposed asymmetry images.

@@

\begin{figure*}
  \vspace{-20mm}
  \makebox[\textwidth][c]{
  \includegraphics[width=140mm]{challengeResults.pdf}
  }
  \caption{Visual results (every tenth slice sampled from the axial center of the
  tumor region) from the BRATS 2013 challenge using the proposed
  framework (off-white = non-enhancing tumor, cyan = edema, violet = enhancing 
  tumor, yellow = necrosis).  Tumor components were combined into
  three regions for performance assessment ((1) complete tumor: all four tissue
  classes; (2) tumor core:  necrosis, non-enhancing tumor, and enhancing tumor;
  and (3) enhancing tumor).  See Table \ref{table:results} for quantitative measures
  and challenge rankings.
  }
  \label{fig:challengeresults}
\end{figure*}

@@

\begin{table*}
\caption{Assessment measures from the MICCAI 2013 BRATS Challenge including
 Dice overlap, positive prediction value (PPV), and sensitivity.}
\label{table:results}
\begin{center}
\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill} } l c c c c}
\toprule
\multicolumn{1}{c}{ \bf Data Set} & {\bf Rank}$^\dagger$ & {\bf Dice} & {\bf PPV} & {\bf Sensitivity} \\
\midrule
Challenge (complete) & 1 & 0.87 (1) & 0.85 (2) & 0.89 (2) \\
Challenge (core) & {} & 0.78 (1) & 0.74 (5) & 0.88 (1) \\
Challenge (enhanced) & {} & 0.74 (1) & 0.69 (4) & 0.83 (1) \\
\midrule
Leaderboard (complete) &  1 & 0.79 (2) & 0.83 (1) & 0.81 (4) \\
Leaderboard (core)     & {} & 0.65 (1) & 0.70 (2) & 0.73 (2) \\
Leaderboard (enhanced) & {} & 0.53 (3) & 0.51 (3) & 0.66 (2) \\
\midrule
Evaluation (complete) &  2 & 0.88 (2) & 0.88 (3) & 0.89 (3) \\
Evaluation (core)     & {} & 0.76 (2) & 0.80 (5) & 0.79 (3) \\
Evaluation (enhanced) & {} & 0.55 (3) & 0.65 (3) & 0.53 (3) \\
\bottomrule
\multicolumn{5}{l}{\scriptsize$^\dagger$Indicates performance
for all measures over all regions.  Component rankings are 
given in parentheses.
}
\end{tabular*}
\end{center}
\end{table*}

@@

%% References with bibTeX database:

%\section*{Acknowledgments}
%\clearpage
\section*{References}

\bibliographystyle{elsarticle-harv}
\bibliography{references}
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}

@@

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}

\usepackage{color,bm,natbib}
\usepackage[colorlinks=true,breaklinks]{hyperref}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{amsmath}
\usepackage{amssymb}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}
}
%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Nicholas J.~Tustison \and 
        K.~L.~Shrinidhi \and
        Max Wintermark \and
        Christopher R.~Durst \and
        Benjamin M.~Kandel \and
        James C.~Gee \and
        Murray C.~Grossman \and
        Brian B.~Avants
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{N. Tustison \at
              Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA \\
%              Tel.: +123-45-678910\\
%              Fax: +123-45-678910\\
              \email{ntustison@virginia.edu}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
              \and
              K. Shrinidhi, B. Kandel, J. Gee, and M. Grossman \at
              Penn Image Computing and Science Laboratory, Department of Radiology, University of Pennsylvania, Philadelphia, PA
              \and
              M. Wintermark, C. Durst \at
              Department of Radiology and Medical Imaging, University of Virginia, Charlottesville, VA \\
              \and
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Segmenting and quantifying gliomas from MRI is an important task for diagnosis, planning intervention, 
and for tracking tumor changes over time.
However, this task is complicated by the lack
of prior knowledge concerning tumor location, spatial extent, shape,
possible displacement of normal tissue, and intensity signature.
%These factors limit the applicability of techniques with established
%performance in related tasks such as normal brain tissue
%segmentation.  
To accommodate such complications, we introduce a framework for supervised
segmentation based on multiple modality intensity, geometry, and asymmetry feature sets.  These features drive a supervised whole-brain and tumor segmentation
approach based on random forest-derived probabilities.  The
asymmetry-related features (based on optimal symmetric multimodal 
templates) demonstrate excellent discriminative 
properties within this framework. We also gain performance by  
generating probability maps from random forest models and using these maps 
for a refining Markov random field regularized probabilistic segmentation.  
This strategy allows us to interface the supervised learning capabilities 
of the random forest model with regularized probabilistic segmentation
using the recently developed \textit{ANTsR} 
package---a comprehensive statistical and visualization interface between the
popular Advanced Normalization Tools (ANTs) and the \textit{R}
statistical project. 
%\textit{ANTsR} integrates proven image analysis methods
%for image registration, segmentation, bias correction, and
%template building with \textit{R}'s statistical and machine
%learning techniques and state-of-the-art visualization possibilities.  
The reported algorithmic framework was the
top-performing entry in the MICCAI 2013 Multimodal Brain Tumor
Segmentation challenge.  The challenge data were widely varying 
consisting of both high-grade and low-grade glioma tumor 
four-modality MRI from five different institutions.
Average Dice overlap measures for the final algorithmic assessment
were 0.87, 0.78, and 0.74 for ``complete'', ``core'', and ``enhanced'' 
tumor components, respectively.
%and, to our knowledge, is the only entry for
%both challenge years (2012 \& 2013) which has been made publicly
%available.
\keywords{advanced normalization tools \and BRATS \and glioma \and \textit{R} project
}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}

Given the complexity of tumor growth and appearance and the need
for precise volumetric measurements for tumor characterization, 
much research has been invested in computational methods for 
automated segmentation of tumor regions in MR images.  Several 
approaches have been previously proposed in the literature as
detailed in recent reviews \citep{angelini2007,bauer2013}.  
Commensurable algorithmic evaluation, however, is
extremely problematic due to widely varying performance assessments
described in the corresponding publications, lack of publicly available 
evaluation data, and private algorithmic instantiations.  
In response to these issues, the Multimodal Brain Tumor Segmentation 
(BRATS) challenge was initiated in 2012 (and continued in 2013) under 
the auspices of the Medical Image Computing and Computer Assisted 
Intervention Society in association with their annual international 
conference \citep{menze2014}.

@@

@@

\begin{figure}
  \centering
  \begin{tabular}{cccc}
  \includegraphics[width=28mm]{Fig1a.eps} &
  \includegraphics[width=28mm]{Fig1b.eps} &
  \includegraphics[width=28mm]{Fig1c.eps} &
  \includegraphics[width=28mm]{Fig1d.eps} \\
  (a) & (b) &
  (c) & (d) \\
  \end{tabular}
  \caption{Induced bilateral asymmetry due to tumor presence causing
  distortion of the plane of symmetry.  Shown are mid-axial slices of
  one of the BRATS 2013 training data 
  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
  (d) T2.  }
  \label{fig:asymmetry}
\end{figure}

\begin{figure*}
    \centerline{\includegraphics[width=120mm]{Fig2.eps}}
  \caption{Multivariate symmetric template created from the MMRR 
           data \citep{landman2011}.  Of the seven modalities 
           comprising the set of study acquisitions, we illustrate the
           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
           Although DWI-based images were not included in the challenge data, 
           such images have shown discriminative potential \citep{price2003,cha2005} 
           warranting investigation in our future work.  
           The optimal transformation and averaging of the individual 
           subject images result in the asymmetric template represented at 
           the top of the middle column.  A horizontal reflection, $T_R$,
           perpendicular to the mid-sagittal plane  resulted in the 
           contralateral counterpart represented
           at the bottom of the second column.  The final template seen on
           the right is a result of repeating the template construction using
           the two asymmetric templates as input.
          }
  \label{fig:symmetrictemplates}
\end{figure*}

@@

\begin{figure}
  \centering
    \includegraphics[width=100mm]{Fig3.eps}
  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.  $T_R$ denotes a horizontal reflection
  perpendicular to the mid-sagittal plane.
  Feature images used are specified by the dashed box.
Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
          }
  \label{fig:asymmetryFeatures}
\end{figure}

@@

\begin{figure*}
\centering
  \makebox[0.9\textwidth][c]{
  \includegraphics[width=120mm]{Fig4.eps}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  final assessment data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Similarly, we 
           calculate the voxelwise difference image between the
           T1 and T1C image.  
           of 
           Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   This mapping is also used to produce the log Jacobian image
           and warped template distance mask.  Additionally, the 
           (T1 - T1C) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

@@

\begin{figure}
  \centering
   \vspace{-15mm}
  \makebox[0.9\textwidth][c]{
    \includegraphics[width=150mm]{Fig5.eps}
    }
  \caption{Diagrammatic workflow for the proposed RF model training (left) and prediction
  (right).  For training, the feature images are first generated from the set of input 
  training multimodal MRI and symmetric multivariate template. The set of features images are 
  use to create the RF model for the first stage.  This first stage RF model is then applied
  to the set of input training data to yield spatial priors for seeding the generation of the
  second stage feature images.  These feature images are then used to create the second
  stage RF model.  For prediction of a single-subject segmentation, a similar scenario 
  is applied whereby the generated feature images are used as input to the RF models the
  second of which produces an initial segmentation estimate.  This estimate is then refined
  using a series of heuristically-derived binary morphological operations resulting in the 
  final labeled brain.
  }
  \label{fig:pipeline}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tabular}{c}
    \includegraphics[width=88mm]{Fig6.eps}
  \end{tabular}
  \caption{Visual comparison between the first and second RF stages
  on the evaluation {\tt BRATS\_HG0001} data set.
  Although the GMM intensity modeling does provide certain 
  discriminative benefit (see Figures \ref{fig:hgimportance}
  and \ref{fig:lgimportance}), the use of the RF-derived posteriors 
  coupled with the described MAP-MRF segmentation framework enhances 
  the accuracy of the intensity-based modeling features.    
  }
  \label{fig:stageComparison}
\end{figure}

@@

Additionally, the {\tt randomForest} package provides  measurements 
for determining the importance of chosen features when constructing the model.  
This aids in potential feature pruning or intuiting model behavior.  As mentioned
earlier, we provide the \textit{R} script {\tt plotVariableImportance.R} to render
one such quantity use to assess RF importance using {\tt MeanDecreaseAccuracy}.  
During model construction
(specifically the out-of-bag error calculation stage), the decrease in prediction accuracy
with the omission of a single feature or variable is tracked and averaged.  Thus,
those features which have the greatest decrease in mean accuracy are considered
to be the most discriminative.
%\footnote{
%There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.  Correctives have been proposed but we ignore these issues for this particular application.
%}
In this work, we do not use these measurements for feature pruning.  However,
we plot them in the Results section (see Figures \ref{fig:hgimportance} and
\ref{fig:lgimportance}) as they demonstrate the relative importance of our
selected features including that of the proposed asymmetry images.

@@

\begin{figure*}
  \vspace{-20mm}
  \makebox[\textwidth][c]{
  \includegraphics[width=140mm]{Fig7.eps}
  }
  \caption{Visual results (every tenth slice sampled from the axial center of the
  tumor region) from the BRATS 2013 challenge using the proposed
  framework (off-white = non-enhancing tumor, cyan = edema, violet = enhancing 
  tumor, yellow = necrosis).  Tumor components were combined into
  three regions for performance assessment ((1) complete tumor: all four tissue
  classes; (2) tumor core:  necrosis, non-enhancing tumor, and enhancing tumor;
  and (3) enhancing tumor).  See Table \ref{table:results} for quantitative measures
  and challenge rankings.
  }
  \label{fig:challengeresults}
\end{figure*}

@@

\begin{table*}
\caption{Assessment measures from the MICCAI 2013 BRATS Challenge including
 Dice overlap, positive prediction value (PPV), and sensitivity.}
\label{table:results}
\begin{center}
\begin{tabular*}{0.99\textwidth}{@{\extracolsep{\fill} } l c c c c}
\toprule
\multicolumn{1}{c}{ \bf Data Set} & {\bf Rank}$^\dagger$ & {\bf Dice} & {\bf PPV} & {\bf Sensitivity} \\
\midrule
Challenge (complete) & 1 & 0.87 (1) & 0.85 (2) & 0.89 (2) \\
Challenge (core) & {} & 0.78 (1) & 0.74 (5) & 0.88 (1) \\
Challenge (enhanced) & {} & 0.74 (1) & 0.69 (4) & 0.83 (1) \\
\midrule
Leaderboard (complete) &  1 & 0.79 (2) & 0.83 (1) & 0.81 (4) \\
Leaderboard (core)     & {} & 0.65 (1) & 0.70 (2) & 0.73 (2) \\
Leaderboard (enhanced) & {} & 0.53 (3) & 0.51 (3) & 0.66 (2) \\
\midrule
Evaluation (complete) &  2 & 0.88 (2) & 0.88 (3) & 0.89 (3) \\
Evaluation (core)     & {} & 0.76 (2) & 0.80 (5) & 0.79 (3) \\
Evaluation (enhanced) & {} & 0.55 (3) & 0.65 (3) & 0.53 (3) \\
\bottomrule
\multicolumn{5}{l}{\scriptsize$^\dagger$Indicates total performance
 over all regions.  Component rankings are 
in parentheses.
}
\end{tabular*}
\end{center}
\end{table*}

@@

\section*{Disclosures}
The authors declare that they have no conflict of interest.

%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

% BibTeX users please use one of
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base

\bibliographystyle{spbasic}      % basic style, author-year citations
\bibliography{references}

\end{document}
% end of file template.tex
